---
author: "Ziyun Ding"
output:
  html_document: default
  pdf_document: default
  word_document: default
---
###1. Data exploration.
```{r}
library(ggplot2)
library(tidyverse)
```

```{r}
data <-read.csv("forestfires.csv",sep=',',header=TRUE)
data<-as.data.frame(data)
attach(data)
```
```{r}
#Figure 1. plot distribution of area
ggplot(data, aes(x = area))+geom_histogram(aes(y = ..density..), fill = "light blue", colour = "black", bins = 40)
```

```{r}
#Figure 2. plot distribution of the fire
data %>% group_by(X, Y) %>% summarise(n=n()) %>% ggplot(aes(x=X, y=Y, size = n, color = n)) + geom_point() + scale_x_continuous(breaks = seq(1, 9, 1)) + scale_y_continuous(breaks = seq(1, 9, 1)) + xlab('X coordinate') + ylab('Y coordinate') + scale_colour_gradient(low = "light blue", high = "dark blue")
```

```{r}
#count the number of burned area and not burned area
round(table(area==0), 2)
```

```{r}
#Table 1. find the region has most fire occurance
data %>% filter(area > 0) %>% group_by(X, Y) %>% summarise(count = n(), max_area = max(area)) %>% arrange(desc(count)) %>% .[1:10,]
```
```{r}
#Table 1. find the region has largest burned area.
data %>% filter(area > 0) %>% group_by(X, Y) %>% summarise(count = n(), max_area = max(area)) %>% arrange(desc(max_area)) %>% .[1:10,]
```

###2. Factors Influencing Fire Occurrence

By creating a dummy variable, factor_area with tow levels (factor_area=1 means the area was bured while factor_area=0 means the area wasn't burned), we are able to analysis the probability of the occurrance of fire with a binomial model.

1.1 Data Transformation and Model Selection
```{r}
databi<-as.data.frame(data)
databi$factor_area <- rep(0, 517)
databi$factor_area[databi$area>0] <- 1
```

Because there are data unavailable in some levels of categorical factors month, we transform the month factor into the season factor. 

```{r}
#we re-classify the month factor into different seasons
databi$season <- rep("spring", 517)
for (i in 1:517){
if (databi$month[i] %in% c("feb","jan","dec")) databi$season[i] <- "winter"
if (databi$month[i] %in% c("oct","nov","sep")) databi$season[i] <- "autumn"
if (databi$month[i] %in% c("aug","jul","jun")) databi$season[i] <- "summer"
}
databi$season <- as.factor(databi$season)
databi$count<- rep(1, 517)
(table<-xtabs(count~ season+factor_area, databi))
```

```{r}
#refit model with season
bmod3<-glm(factor_area~season+(FFMC+DMC+DC+ISI+temp+RH+wind)^2+I(FFMC^2)+I(DMC^2)+I(DC^2)+I(ISI^2)+I(temp^2)+I(RH^2)+I(wind^2),data=databi,family=binomial(link=logit))
bmod4<-step(bmod3,test="Chi", trace =FALSE)
summary(bmod4)
```
```{r}
#estimate of the overdispersion parameter
sum(residuals(bmod4,type="pearson")^2)/df.residual(bmod4)
```

1.2 Model Diagnosis
```{r}
#Figure 3. Diagnosis plots for the logistic regression model 
par(mfrow=c(1,3))
plot(bmod4, which=c(1,2, 4))
```


1.3 5-fold Cross Validation
```{r}
##cross validation using logistic regression model, using the binary response (burned or not) as response
#fold <- 5
#cut_off <- 0.7
crossval <- function(fold, cut_off){
  for(i in seq(floor(nrow(databi)/fold), nrow(databi), by=floor(nrow(databi)/fold))){
          testingX <- databi[(i-floor(nrow(databi)/fold)+1):i, 5:12]
          testingy <- databi[(i-floor(nrow(databi)/fold)+1):i, 13]
          testing <- list(testingX, testingy)
          training <- databi[-((i-floor(nrow(databi)/fold)+1):i),]
          #trainingX <- databi[-(i:(i+fold-1)), 5:12]
          #trainingy <- databi[-(i:(i+fold-1)), 13]
          #training <- list(trainingX, trainingy)
          log_pred <- predict(bmod4, testingX)
          pred_prob <- exp(log_pred)/(1+exp(log_pred))
          pred_prob[pred_prob > cut_off] <- 1
          pred_prob[pred_prob < cut_off] <- 0
          print(sum(pred_prob==testingy)/length(testingy))
  }
}
```

```{r}
crossval(5, 0.8)
```

## Part 2: Factors Influencing Burned Area

2.1 Data Transformation
Based on the histogram, we can see that the distribution of burned area is highly skewed. Hence, the Y (area) need to be transformed in order to continue with the analysis.
```{r}
datamul <- data[area>0,]
#Figure 4. plot distribution of area >0
ggplot(datamul, aes(x = area))+geom_histogram(aes(y = ..density..), fill = "light blue", colour = "black", bins = 40)
```
```{r}
# box-cox transformation suggest using log(area)
lmod1<-lm(area ~month+day+(FFMC+DMC+DC+ISI+temp+RH+wind)^2+I(FFMC^2)+I(DMC^2)+I(DC^2)+I(ISI^2)+I(temp^2)+I(RH^2)+I(wind^2), data = datamul)
library(car)
boxCox(lmod1,lamda= seq(-2, 2, 1/10))
```
```{r}
#Figure 5. plot area after transformation
ggplot(datamul, aes(x = log(area)))+geom_histogram(aes(y = ..density..), fill = "light blue", colour = "black", bins = 40)+stat_function(fun = dnorm, args = list(mean = mean(log(area[area!=0])), sd = sd(log(area[area!=0]))), colour = 'orange')
```


2.2 Model Selection
```{r}
# model selection
lmod2<-lm(log(area) ~month+day+(FFMC+DMC+DC+ISI+temp+RH+wind)^2+I(FFMC^2)+I(DMC^2)+I(DC^2)+I(ISI^2)+I(temp^2)+I(RH^2)+I(wind^2), data = datamul)
lmod3<-step(lmod2,test="Chi",trace=F)
summary(lmod3)
```

2.3 Model Diagnostics
```{r}
#Figure 6. Diagnosis plot for the linear regression model (2).
par(mfrow=c(2,2))
plot(lmod3, which=c(1,2, 3, 4))

```

2.4 5-fold Cross Validation
```{r}
##cross validation using linear regression model, using the burned area as response
fold <- 5
acc <- rep(0, fold)
for(i in seq(floor(nrow(datamul)/fold), nrow(datamul), by=floor(nrow(datamul)/fold))){
  correct_count <- 0
  testingX <- datamul[(i-floor(nrow(datamul)/fold)+1):i, 2:12]
  testingy <- datamul[(i-floor(nrow(datamul)/fold)+1):i, 13]
  testing <- list(testingX, testingy)
  training <- datamul[-((i-floor(nrow(datamul)/fold)+1):i), 2:13]

  pred_res <- predict(lmod3,testingX,interval="confidence")
  for (i in 1:length(testingy)){
    if ((log(testingy[i]) >= pred_res[i,2]) & (log(testingy[i]) <= pred_res[i,3])){
      correct_count=correct_count+1
    }
  }
  print(correct_count/length(testingy))
}
```


